---
title: "HW Class 17: UNIX Basics"
author: Christopher Brockie (PID A16280405)
format: pdf
---

# Using RStudio to read your output

Read the zebrafish output file:
```{r}
zebrafish <- read.table("mm-second.x.zebrafish.tsv", col.names = c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore"))
```

Check to ensure table has been read correctly:
```{r}
head(zebrafish)
```

Make a histogram of the `bitscore` values:
```{r}
hist(zebrafish$bitscore, breaks = 30, xlab = "Bitscore", ylab = "Frequency", main = "Histogram of Bitscore Values")
```
Most of the `bitscore` values are on the lower side.

`bitscores` are only somewhat related to `pident`; they take into account not only the percent identity but the length of the alignment. This relationship can be estimated by the following napkin sketches:
```{r}
library(ggplot2)
ggplot(zebrafish, aes(pident, bitscore,)) + 
  geom_point(alpha=0.1) +
  labs(x="Percent Identity", y="Bitscore")
```

```{r}
ggplot(zebrafish, aes((pident * (qend - qstart)), bitscore)) + 
  geom_point(alpha=0.1) + 
  geom_smooth() +
  labs(x="Percent Identity x Length", y="Bitscore")
```


