---
title: "Class 06: R Functions"
author: "Christopher Brockie (PID: A16280405)"
format: pdf
---

# All about functions in R

Functions are the way we get stuff done in R. We call a function to read data, compute stuff, plot stuff, etc.

R makes writing functions accessible but we should always start by trying to get a working snippet of code first before we write our function.

## Today's Lab

We will grade a whole class of student assignments. We will always try to start with a simplified version of the new problem.

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

If we want the average we can use the `mean()` function
```{r}
mean(student1)
```

Let's be nice instructors and drop the lowest score so the answer here should be 100. 

I can use the `min()` function to find the lowest value.

```{r}
min(student1)
```

I found the `which.min()` function that may be useful here. How does it work? Let's just try it:

```{r}
student1
which.min(student1)
```

I can use the minus syntax trick to get everything but the element with the min value.

```{r}
student1[-which.min(student1)]
```

I have my first working snippet of code :-)
```{r}
mean(student1[-which.min(student1)])
```

Let's test on the other students
```{r}
student2
mean(student2[-which.min(student2)])
```

Where is the problem - oh it is the `mean()` with NA input returns NA without including syntax `na.rm()` in the `mean()` argument.
```{r}
mean(student2, na.rm = TRUE)

```

Lets test on student 3
```{r}
mean(student3, na.rm=TRUE)
```

No bueno. We need to fix this!

I want to stop working with `student1`, `student2`, `student3` etc. and typing it out everytime so lets instead work with an input called `x`.

```{r}
x <- student2
x
```

We want to overwrite the NA values with zero - if you miss a homework you score zero on this homework.

Google and Claude told me about the `is.na` function. Lets see how it works.

```{r}
x
is.na(x)
```

```{r}
x[is.na(x)] <- 0
x
```

We can use logicals to index a vector.

```{r}
y <- 1:5
y
y > 3
y[y > 3]
y[y >3] <- 100
```

This is my working snippet of code that solves the problem for all of my example inputs :-)

```{r}
x <-  student3
# Mask NA values to zero
x[is.na(x)] <- 0
#Drop lowest score and get mean
mean(x[-which.min(x)])
```

> Q1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
grade <- function(x) {
  # Mask NA values to zero
  x[is.na(x)] <- 0
  #Drop lowest score and get mean
  mean(x[-which.min(x)])
}
```

Use this function:
```{r}
grade(student1)
grade(student2)
grade(student3)
```

We need to read the gradebook. We use `row.names()` in order to make the students the primary column.

```{r}
gradebook <- read.csv("https://tinyurl.com/gradeinput",
                      row.names=1)
gradebook
```

I can use the `apply` function if I figure out how to use it:

```{r}
ans <- apply(gradebook, 1, grade)
ans
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

```{r}
which.max(ans)
```

> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

We could calculate the `mean()` score for each homework.

```{r}
mask <- gradebook
mask[is.na(mask)] <- 0
hw.ave <- apply(mask, 2, mean)
hw.ave
which.min(hw.ave)
```

We could take the `sum()`.

```{r}
hw.sum <- apply(gradebook, 2, sum, na.rm=T)
which.min(hw.sum)
```

> Q4. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
apply(mask, 2, cor, y=ans)
```

```{r}
which.max(apply(mask, 2, cor, y=ans))
```

